Simultaneous Settler Projects and U.S. Racial State Formation

### 2024-06-28 ms_backend.Rmd
fast forward before merge with front end

```{r read, include = F, message = F}
### state shape file
st<-states(cb = T) %>%
  st_transform(8528) %>% 
  filter(!STUSPS%in%
           c("HI", "AK", 
             "PR", "GU", "AS","VI",
             "MP"))
# read data from make_* scripts
pop<-read_csv("./data/temp_population_ts.csv")%>%
  filter(STUSPS!="AK")
disp<-read_csv("./data/temp_disposession_ts.csv")%>%
  filter(STUSPS!="AK")
dist<-read_csv("./data/temp_distribution_ts.csv")%>%
  filter(STUSPS!="AK")
bs<-read_csv("./data/temp_bs_ts.csv")
agents<-read_csv("./data/temp_agents.csv")
## add region to all
disp<-disp %>%
  left_join(pop %>%
              select(STUSPS, region, division) %>%
  distinct()) %>% 
  left_join(bs) %>% 
  left_join(agents %>% 
              select(-state))
dist<-dist %>%
  left_join(pop %>%
              select(STUSPS, region, division) %>%
  distinct()) %>% 
  left_join(bs) %>% 
  left_join(agents %>% 
              select(-state))
```

```{r run-models, cache = T, include = F, message = F}
# source("models.R")
mdat<-pop %>% 
  left_join(dist) %>% 
  left_join(bs) %>% 
  mutate(blk_sharecrop_pc = blk_sharecrop / tot_pop,
         urban_immig_pc = urban_immig_pop / tot_pop,
         homestead_plus_morill = homestead_acres_cum + morill_acres_cum)

#### set weakly informative priors for model parameters
### for m1
### try a growth model
mdat<-mdat %>% 
  mutate(year_c = (year - 1900)/10)

teachers_m1<-readRDS("./models/teachers_m1.rds")
cops_m1<-readRDS("./models/cops_m1.rds")
socwork_m1<-readRDS("./models/socwork_m1.rds")
prison_m1<-readRDS("./models/prison_m1.rds")
inst_m1<-readRDS("./models/inst_m1.rds")
bs_m1_z<-readRDS("./models/bs_m1_z.rds")
bs_m1<-readRDS("./models/bs_m1.rds")
```

@@

We describe the relationship between settler colonial political and economic contexts and state formation by evaluating whether a set of measures of policy infrastructure are related to a set of measures that are related to locally-specific settler colonial policy contexts at the state and year-levels. We use data on US states between 1900 and 1930 derived from US Census and archival sources.

Our approach is theory-driven, informed by 1) RACIAL STATE THEORY and 2) the HISTORIOGRAPHY OF EACH REGION IN THIS PERIOD. We suggest that RACIAL STATE THEORY provides a coherent framework for understanding heterogeneity in state formation during this period, by applying the simultaneous settler projects framework to these distinct contexts.

@@

*While there have been a number of compelling analyses of the long-run effects of variation in colonial systems on colonized societies (ACEMOGLU ETC) and of chattel slavery and racial terror on contemporary politics (CITATIONS) we refrain from engaging in formal causal inference in this analysis. Rich theoretical and historical work suggests that racial and settler colonial processes deeply structure policy outcomes through complex and long-running processes that become obscured and entangled with a battery of social processes. We suggest that these processes are visible at a relatively high-level in the historical data. However, settler colonialism is endogenous to the whole of US politics, economics, and demography.* TIGHTEN THIS

@@

Our first set of models (see equation XXX) predict infrastructure outcomes for each decennial Census year between 1900 and 1930, and include time-stable state-level random intercepts, a non-linear time component (a national decade-specific intercept), and parameters for per capita rates of Black tenant farming, per capita rates of white urban immigrants, and cumulative acreage redistributed to settlers under the Homestead and Morill acts. We use Bayesian Hamiltonian Monte Carlo methods to estimate all models with weakly informative priors on all model parameters. This approach adjusts for national time trends (secular increases in employment across places), as well as adjusting for idiosyncratic differences across states.

@@

$y \sim \textrm{NegBin}(\lambda, \alpha)$

$\lambda = \textrm{E}[y]$

$\log \left( \frac{\textrm{E}[y]}{\textrm{population}} \right) = \gamma_s + \beta \textrm{X}_{st}$

$x_{st} = t \times \left(\frac{\textrm{Black tenant farmers}_{st}}{\textrm{population}_{st}} +  \frac{\textrm{White urban immigrants}_{st}}{\textrm{population}_{st}} +\textrm{Morill and Homestead acres}_{st}\right)$

$\gamma \sim N(0, \sigma^2_\gamma)$

@@

$\beta \sim N(0, 20)$

$\sigma \sim t(3, 0, 2.5)$

$\alpha \sim \textrm{Gamma}(0.01, 0.01)$

@@

$ y > 0 \sim \textrm{Bernoulli}\left(\theta\right) $
$ \textrm{logit} (\theta) =  \beta_{0t} + \beta_1 \frac{\textrm{Black tenant farmers}_{s}}{\textrm{population}_{s}} + \beta_2 \frac{\textrm{White urban immigrants}_{s}}{\textrm{population}_{s}} + \beta_3 \textrm{Morill and Homestead acres}_{s} $
$\beta \sim N(0, 20)$

For all regression models, we transform predictors to a mean-centered standard deviation scale (*z* scores) to improve model convergence in a context of widely differing scales across measures.

@@

The extensive and methodical dispossession of Native lands that characterized the Assimilation and Allotment era was in full force during the Progressive Era. Panel 1 of Figure \@ref(fig:three-maps) shows the cumulative area of dispossessed Native lands that had been redistributed to settlers and US states at low or no cost under the Homestead and Morill Acts of 1862 by 1900. Nearly 90 million acres of Native lands had been granted to settlers and states through the Morill and Homestead acts by 1900. By 1930, this total would reach over 265 million acres. By 1930, 47 percent of the contemporary area of Nebraska, and 44 percent of the contemporary area of North Dakota had been granted to settlers under these programs. More than 30 percent of the contemporary area of South Dakota, Oklahoma, Colorado, and Montana had been granted to settlers under these programs by 1930. These programs primarily impacted lands in the US West, Southwest, Mountain West, and Midwest during this period.

@@

The Jim Crow system of apartheid and its economic complement the sharecropping system were developed and became deeply rooted across the states in the US South during the Progressive era. Panel 2 of Figure \@ref(fig:three-maps) shows the entrenchment of the sharecropping system in 1930, which was still a dominant mode of racialized political economy in the US South several decades into the first Great Migration of African Americans out of the US South to industrial cities in the Northeast and Midwest. In 1930, 644,000 Black residents of Mississippi were living on tenant farms, 32 percent of the state's total population. More than 400,000 Black people were recorded as residents of tenant famrs in Georgia, Alabama, and South Carolina in 1930. More than 300,000 were enumerated in North Carolina, Louisiana, and Texas.  And more than 100,000 were enumerated in Arkansas, Tennessee, and Virginia.

And the Industrial Revolution, coupled with a wave of crises across Europe, created both strong push and pull forces for large populations of immigrants to travel to North America. Southern and Eastern European immigrants settled across the industrial cities of the Northeast, Midwest, and West coasts. Many of them also joined the waves of white settlers pursuing the opportunity to claim recently dispossessed Native lands in the Central and Western portions of the US. As shown in Panel 3 of Figure \@ref(fig:three-maps), the populations of many states in the Northeast, Midwest, and West Coast had become increasingly urban and foreign born. Nearly 3 million residents of New York state's 12.5 million residents in 1930 were either first or second generation European immigrants. In Illinois, 1.1 million of the state's 7.6 million residents were first or second generation immigrants from Europe. In Massachusetts and Rhode Island, 23 percent of the each state's population were European immigrants in 1930. More than 10 percent of the population were European immigrants in New Jersey, Connecticut, Michigan, New Hampshire, California, and Pennsylvania.

@@

We define 'social state capacity' as the resources available to governments to direct and intervene in processes of social control and social reproduction (see Cingolani 2013 for a review of theory and measurement of the more general 'state capacity'). Capacity is a function of both physical infrastructure and human resources available to state actors. During the Progressive Era, federal, state, and local governments developed their abilities to intervene into and manage social life through a series of increasingly professionalized bureaucracies. We suggest that these nascent components of the carceral and welfare states developed in markedly different ways across places. To capture some of this heterogeneity, we focus our analysis on a employment and infrastructure outcomes in a broad set of social policy systems including education, social welfare, and criminal justice.

@@

Employment rates for teachers were highest in a cluster of Midwestern and Mountain West states. Thirteen states had teacher employment rates of over 10 per 1,000 population, all in the Midwest or West. North and South Dakota had highest-in-the-nation teacher employment rates, at over 14 per 1,000 persons. Teachers were instrumental in building human capital and national culture in settler homesteading communities. They also served as front-line agents of the assimilationist programs throughout the region's vast network of Indian boarding schools (Marquez 2024, OTHERS). Employment rates for teachers were lowest-in-the nation in the US South. There were more than twice as many teachers per capita in North Dakota in 1930 than there were in Alabama. In Alabama, Arkansas, Kentucky, Missisippi, and Georgia, the 1930 Census recorded fewer thatn 7 teacher per 1,000 residents. Southern states in general systematically underfunded or avoided funding public institutions that could benefit African Americans, and the Jim Crow school system was an exemplar of this practice (CITE).

@@

US welfare and carceral systems have historically prioritized developing physical infrastructure to confine and/or treat populations understood to be socially problematic (CITES ON CONFINEMENT). In Figure \@ref(fig:gq-map) we show the distribution of the population housed in institutional group quarters in 1930 across US states as recorded by the US Census to provide one measure of physical infrastructure for social welfare and social control. We display the population housed in prisons in jails in panel 1, the population housed in mental institutions in panel 2, and the population housed in institutions for the poor and/or disabled in figure 3. Note that data quality issues in the 1910 Census limit the possibilities for time series analysis.

@@

Panel 1 of Figure \@ref{gq-map} shows state-level variation in confinement in correctional group quarters reported in the 1930 US Census. We find average confinement rates to be lowest in the Mid-Atlantic and New England states, at 1.2 per 1,000 population. Confinement rates were highest on average in the West Coast states (2.5 per 1,000). Average rates across the remaining regions varied between 1.3 and 2.0 per 1,000 population. Kansas, home to federal and military prisons in Leavenworth, had the highest incarceration rates in the nation, at 4.5 persons per 1,000 population.

@@

Indian boarding schools were built by the US government and its allies in to eliminate Native nations through the coercive assimilation of Indigenous children (Adams 1995; Hoxie 2001). Investment in Indian boarding schools peaked during the Assimilation and Allotment era of Federal Indian Law, roughly spanning 1887 - 1934. Notably, these carceral and genocidal institutions represented one of the earliest sustained investments by the federal government in education infrastructure. Below, we provide a descriptive overview of the geography and timing of boarding school operations. Following initiatives by Secretary of the Interior Deb Haaland and leaders from across Indian Country, the US Department of the Interior has initiated its first comprehensive review of the scope and impacts of the Indian boarding school program. We rely on data produced by the Department of the Interior (NEWLAND 2022) for historical counts of boarding schools in operation by state and year.

Figure \@ref(fig:bsts) displays the number of Indian boarding schools operating annually between 1830 and 1930, with eras of Federal Indian law shaded for reference. The number of boarding schools grew dramatically during the Reservation and Assimilation eras, with most growth occurring between 1870 and 1905. After 1905, a slow decline in the number of operating schools began and continued until the end of our focal period, 1930. At the beginning of the Removal Era (1830), there were 18 boarding schools in operation, and 23 were operating at the end of the Removal era (1850). During the subsequent Reservation era (1850 - 1887), the number of operating boarding schools increased from 23 to 173. The number of operating facilities continued to increase during the Assimilation era to a maximum of 242 in 1908, followed by a decline to 174 facilities recorded as operating in 1930, a similar level of facilities operating as had existed at the beginning of the period.

@@

Figure \@ref(fig:bsmap) shows the geographic distribution of boarding schools in operation when enrollments were at a 1830 - 1930 period maximum (242 facilities documented as operating in 1907). The Department of the Interior has identified Indian Boarding schools operating in 1907 in most regions of the country, with the exception of New England and the Southeast. There were at least 64 boarding school facilities operating in Oklahama in 1907, by far the highest number in the US. There were 8 states with at least 10 facilities in operation in 1907: South Dakota (24), Minnesota (18), Arizona (14), Montana (14), New Mexico (11), Washington (11), Alaska (10, not displayed), and Wisconsin (10). There were at least 5 baording schools in operation in 7 states: California (7), North Dakota (7), Oregon (7), Colorado (5), Hawaii (5, not displayed), Idaho (5), and Kansas (5). States with at least one confirmed facility in operation in 1907 included Nebraska (3), Nevada (3), Pennsylvania (3), Utah (3), Wyoming (3), Iowa (2), Michigan (2), Arkansas (1), Illinois (1), Missouri (1), New York (1), North Carolina (1), and Virginia (1). The data do not record facilities operating in 1907 in the remaining 21 states.

```{r bsmap, fig.cap = "Boarding schools in operation in 1907 by state"}
bs_ts<-agents %>% 
  left_join(bs) %>% 
  left_join(st)
ggplot(bs_ts %>% 
         filter(year == 1907),
       aes(fill = boarding_schools,
           geometry = geometry)) + 
  geom_sf() + 
  scale_fill_viridis_c(trans = "sqrt") + 
  theme_void() + 
  labs(fill = "Facilities\noperating") + 
  theme(legend.position = "bottom")
```

@@

In Figure \@ref(fig:emp-m-vis) we display our estimates of the spatial and temporal relationships between employment in key social policy professions and settler colonial project indicators. In Figure \@ref(fig:gq-m-vis) we show our estimates for the relationship between some of the physical infrastructure of the carceral and welfare state and settler colonial projects. And in Figure \@ref(fig:bs-m-mvs) we show our estimates of the spatial and temporal relationships between the locations of Federal Indian boarding schools and these measures of settler colonial project timing and location.

@@

PUNCHY HIGH LEVEL SUMMARY. WHAT DO WE CALL THIS? FRONTIER DISPOSSESSION + NATION BUILDING, URBAN REGULATION + IMMIG INCORPORATION, JIM CROW HYBRID FORMAL+INFORMAL RACIAL TERROR + MALIGN NEGLECT.

## Employment models

As shown in Figure \@ref(fig:emp-ts), there was a dramatic increase in the number of people employed as teachers between 1900, when there were `r emp_pdat %>% filter(year==1900, name=="Teachers") %>% select(value) %>% slice(1)` teachers recorded in the US Census, and 1930, when there was a reported total of `r emp_pdat %>% filter(year==1930, name=="Teachers") %>% select(value) %>% slice(1)` teachers. This amounts to a `r round((emp_pdat_prop %>% filter(year==1930, name=="Teachers") %>% select(prop_change) %>% slice(1)-1)*100)` percent growth in the per capita rate of teacher employment over a 30 year period. We model state-year teacher employment rates using the model described in Equation 1, and display the posterior distributions of focal parameters in Figure \@ref(fig:emp-m-vis) to evaluate how the geography of settler colonial projects related to patterns of teacher employment across places and over time. More details on model parameters and diagnostics are provided in Appendix Table (XXXXXX, full regression table).

Panel 1 in Figure \@ref(fig:emp-m-vis) shows the distributions of the posterior parameter estimates of our relationship between three focal settler colonial contexts and teacher employment. Conditional on our data and model specifications, we find that states with higher levels of settler land redistribution under the Homestead and Morill acts employed more teachers per capita than did other states between 1900 and 1930. We find that states with higher levels of Black tenant farming employed fewer teachers per capita than did other states. States with high urban white immigrant populations employed teachers at rates near the national average.

Panel 2 shows the distributions of focal posterior parameter estimates from our models of police employment. We find that states with higher levels of settler land distribution through the Morill and Homestead acts had lower levels of police employment on average. States with large urban white immigrant populations had significantly higher levels of police employment than did other states during this period. And states with large numbers of Black tenant farmers tended to have lower levels of police employment. Despite relatively low levels of police employment in many Southern states, there were incredible levels of both formal and informal white supremacist violence during this period (CITATIONS).

>>>>

```{r read, include = F, message = F}
### state shape file
st<-states(cb = T) %>%
  st_transform(8528) %>% 
  filter(!STUSPS%in%
           c("HI", "AK", 
             "PR", "GU", "AS","VI",
             "MP"))
# read data from make_* scripts
pop<-read_csv("./data/population_ts_full.csv")%>%
  filter(STUSPS!="AK")
disp<-read_csv("./data/temp_disposession_ts.csv")%>%
  filter(STUSPS!="AK")
dist<-read_csv("./data/temp_distribution_ts.csv")%>%
  filter(STUSPS!="AK")
bs<-read_csv("./data/temp_bs_ts.csv")
## add region to all
disp<-disp %>%
  left_join(pop %>%
              select(STUSPS, region, division) %>%
  distinct()) %>% 
  left_join(bs) 
dist<-dist %>%
  left_join(pop %>%
              select(STUSPS, region, division) %>%
  distinct()) %>% 
  left_join(bs) 
```

```{r run-models, cache = T, include = F, message = F}

mdat<-pop %>% 
  left_join(dist) %>% 
  left_join(bs) %>% 
  mutate(blk_sharecrop_pc = blk_sharecrop / tot_pop,
         urban_immig_pc = urban_immig_pop / tot_pop,
         homestead_plus_morill = homestead_acres_cum + morill_acres_cum)

# rescale year
mdat<-mdat %>% 
  mutate(year_c = (year - 1900)/10)

# source("models.R")

teachers_m1<-readRDS("./models/teachers_m1.rds")
cops_m1<-readRDS("./models/cops_m1.rds")
socwork_m1<-readRDS("./models/socwork_m1.rds")
prison_m1<-readRDS("./models/prison_m1.rds")
mental_m1<-readRDS("./models/inst_m1.rds")
poor_m1<-readRDS("./models/poor_m1.rds")
bs_m1_z<-readRDS("./models/bs_m1_z.rds")
```

@@

We describe the relationship between settler colonial political and economic contexts and state formation by evaluating whether a set of measures capturing social policy infrastructure are related to a set of measures that are related to locally-specific settler colonial policy contexts at the state and year-levels. We use data on US states between 1900 and 1930 derived from US Census and archival sources.

While there have been a number of compelling analyses of the long-run effects of variation in colonial systems on colonized societies (ACEMOGLU ETC) and of chattel slavery and racial terror on contemporary politics (ACHARYA) we refrain from engaging in formal causal inference in this analysis. Rich theoretical and historical work suggests that racial and settler colonial processes deeply structure policy outcomes through complex and long-running processes that become obscured and entangled with a battery of social processes. We suggest that these processes are visible at a relatively high-level in the historical data. Settler colonialism is endogenous to the whole of US politics, economics, and demography. Our interest here is not to identify causal effects, but rather to descriptively chart how the historical geography of settler colonialism in the US overlaps with the historical geography of US state formation. Typical approaches in causal inference seek to eliminate stable cross-sectional differences in units. Our approach prioritizes understanding both these stable cross-sectional differences and dynamics of change over time. Toward this end, we pursue a descriptive and interpretive macro-sociological approach to understanding how settler contexts related to heterogeneity in state formation in the early 20th century.

Our approach is theory-driven, informed by insights derived from 1) RACIAL STATE and SETTLER COLONIAL THEORIES OF US STATE FORMATION and 2) the HISTORIOGRAPHY OF EACH REGION IN THIS PERIOD. We suggest that RACIAL STATE THEORY provides a coherent framework for understanding heterogeneity in state formation during this period, by applying the simultaneous settler projects framework to these distinct contexts.

@@

@@

Our first set of models (see equation XXX) predict infrastructure outcomes for each decennial Census year between 1900 and 1930, and include time-stable state-level random intercepts, a non-linear time component (a national decade-specific intercept), and parameters for per capita rates of Black tenant farming, per capita rates of white urban immigrants, and cumulative acreage redistributed to settlers under the Homestead and Morill acts. We use Bayesian Hamiltonian Monte Carlo methods to estimate all models with weakly informative priors on all model parameters. This approach adjusts for national time trends (secular increases in employment across places), as well as adjusting for idiosyncratic differences across states. Note that the log link function of our negative binomial likelihoods implies multiplicative relationships between additive focal predictors and outcomes, resulting in rates of linear change that vary depending on starting values (intercepts).

@@

$$y \sim \textrm{NegBin}(\lambda, \alpha)$$

$$\lambda = \textrm{E}[y]$$

$$\log \left( \frac{\textrm{E}[y]}{\textrm{population}} \right) = \gamma_s + \beta \textrm{X}_{st}$$

$$x_{st} = t \times \left(\frac{\textrm{Black tenant farmers}_{st}}{\textrm{population}_{st}} +  \frac{\textrm{White urban immigrants}_{st}}{\textrm{population}_{st}} +\textrm{Morill and Homestead acres}_{st}\right)$$

$$\gamma \sim N(0, \sigma^2_\gamma)$$

@@

$$\beta \sim N(0, 20)$$

$$\sigma \sim t(3, 0, 2.5)$$

$$\alpha \sim \textrm{Gamma}(0.01, 0.01)$$

@@

$$ y > 0 \sim \textrm{Bernoulli}\left(\theta\right) $$
$$ \textrm{logit} (\theta) =  \beta_{0t} + \beta_1 \frac{\textrm{Black tenant farmers}_{s}}{\textrm{population}_{s}} + \beta_2 \frac{\textrm{White urban immigrants}_{s}}{\textrm{population}_{s}} + \beta_3 \textrm{Morill and Homestead acres}_{s} $$
$$\beta \sim N(0, 20)$$

For all regression models, we transform predictors to a mean-centered standard deviation scale (*z* scores) to improve model convergence in a context of widely differing scales across measures. More details on model selection are provided in the supplemental materials.

@@

The extensive and methodical dispossession of Native lands that characterized the Assimilation and Allotment era was in full force during the Progressive Era. Panel 1 of Figure \@ref(fig:three-maps) shows the cumulative area of dispossessed Native lands that had been redistributed to settlers and US states at low or no cost under the Homestead and Morill Acts of 1862 by 1930. Nearly 90 million acres of Native lands had been granted to settlers and states through the Morill and Homestead acts by 1900. By 1930, this total would reach over 265 million acres. By 1930, 47 percent of the contemporary area of Nebraska, and 44 percent of the contemporary area of North Dakota had been granted to settlers under these programs. More than 30 percent of the contemporary area of South Dakota, Oklahoma, Colorado, and Montana had been granted to settlers under these programs by 1930. These programs primarily impacted lands in the US West, Southwest, Mountain West, and Midwest during this period.

@@

The Jim Crow system of apartheid and its economic complement the sharecropping system were deeply rooted across the states in the US South during the Progressive era. Panel 2 of Figure \@ref(fig:three-maps) shows the entrenchment of the sharecropping system in 1930, a dominant mode of racialized political economy in the US South several decades into the first Great Migration of African Americans out of the US South to industrial cities in the Northeast and Midwest. In 1930, 644,000 Black residents of Mississippi were living on tenant farms, 32 percent of the state's total population. More than 400,000 Black people were recorded as residents of tenant farms in Georgia, Alabama, and South Carolina in 1930. More than 300,000 were enumerated in North Carolina, Louisiana, and Texas.  And more than 100,000 were enumerated in Arkansas, Tennessee, and Virginia.

And the Industrial Revolution, coupled with a wave of crises across Europe, created both strong push and pull forces for large populations of immigrants to travel to North America. Southern and Eastern European immigrants settled across the industrial cities of the Northeast, Midwest, and West coasts. Many of them also joined the waves of white settlers pursuing the opportunity to claim recently dispossessed Native lands in the Central and Western portions of the US. As shown in Panel 3 of Figure \@ref(fig:three-maps), the populations of many states in the Northeast, Midwest, and West Coast had become increasingly urban and foreign born. Nearly 3 million residents of New York state's 12.5 million residents in 1930 were either first or second generation European immigrants (CONFIRM THIS IS URBAN). In Illinois, 1.1 million of the state's 7.6 million residents were first or second generation immigrants from Europe. In Massachusetts and Rhode Island, 23 percent of the each state's population were European immigrants in 1930. More than 10 percent of the population were European immigrants in New Jersey, Connecticut, Michigan, New Hampshire, California, and Pennsylvania.

@@

We define 'social state capacity' as the resources available to governments to direct and intervene in processes of social control and social reproduction (see Cingolani 2013 for a review of theory and measurement of the more general concept 'state capacity'). Capacity is a function of both physical infrastructure and human resources available to state actors. During the Progressive Era, federal, state, and local governments developed their abilities to intervene into and manage social life through a series of increasingly professionalized bureaucracies. We suggest that these nascent components of the carceral and welfare states developed in markedly different ways across places. To capture some of this heterogeneity, we focus our analysis on employment and infrastructure outcomes in a broad set of social policy systems including education, social welfare, and criminal justice.

@@

Employment rates for teachers were highest in a cluster of Midwestern and Mountain West states. Thirteen states had teacher employment rates of over 10 per 1,000 population, all in the Midwest or West. North and South Dakota had highest-in-the-nation teacher employment rates, at over 14 per 1,000 persons. Teachers were instrumental in building human capital and national culture in settler homesteading communities. They also served as front-line %>% of the assimilationist programs throughout the region's vast network of Indian boarding schools (Marquez 2024, OTHERS). Employment rates for teachers were lowest-in-the nation in the US South. There were more than twice as many teachers per capita in North Dakota in 1930 than there were in Alabama. In Alabama, Arkansas, Kentucky, Missisippi, and Georgia, the 1930 Census recorded fewer than 7 teacher per 1,000 residents. Southern states in general systematically underfunded or failed to develop public institutions that could benefit African Americans, and the Jim Crow school system was an exemplar of this practice (CITE).

@@

US welfare and carceral systems have historically prioritized developing physical infrastructure to confine and/or treat populations understood to be socially problematic (CITES ON CONFINEMENT). In Figure \@ref(fig:gq-map) we show the distribution of the population housed in institutional group quarters in 1930 across US states as recorded by the US Census to provide one measure of physical infrastructure for social welfare and social control. We display the population housed in prisons in jails in panel 1, the population housed in mental institutions in panel 2, and the population housed in institutions for the poor and/or disabled in figure 3. Note that alternative measures of group quarters populations are available, but choice of measures does not have a meaningful impact on our substantive conclusions (for a detailed discussion, see Appendix).

@@

Panel 1 of Figure \@ref{fig:gq-map} shows state-level variation in confinement in correctional group quarters reported in the 1930 US Census. We find average confinement rates to be lowest in the Mid-Atlantic and New England states, at 1.2 per 1,000 population. Confinement rates were highest on average in the West Coast states (2.5 per 1,000). Average rates across the remaining regions varied between 1.3 and 2.0 per 1,000 population. Kansas, home to federal and military prisons in Leavenworth, had the highest incarceration rates in the nation, at 4.5 persons per 1,000 population.

@@

Indian boarding schools were built by the US government and its allies in to eliminate Native nations through the coercive assimilation of Indigenous children (Adams 1995; Hoxie 2001). Investment in Indian boarding schools peaked during the Assimilation and Allotment era of Federal Indian Law, roughly spanning 1887 - 1934. Notably, these carceral and genocidal institutions represented one of the earliest sustained investments by the federal government in education infrastructure (alongside the Morill act of 1862). Below, we provide a descriptive overview of the geography and timing of boarding school operations. Following initiatives by Secretary of the Interior Deb Haaland and leaders from across Indian Country, the US Department of the Interior has initiated its first comprehensive review of the scope and impacts of the Indian boarding school program. We rely on data produced by the Department of the Interior (NEWLAND 2022) for historical counts of boarding schools in operation by state and year.

Figure \@ref(fig:bsts) displays the number of Indian boarding schools operating annually between 1830 and 1930, with eras of Federal Indian law shaded for reference. The number of boarding schools grew dramatically during the Reservation and Assimilation eras, with most growth occurring between 1870 and 1905. After 1905, a slow decline in the number of operating schools began and continued until the end of our focal period, 1930. At the beginning of the Removal Era (1830), there were 18 boarding schools in operation, and 23 were operating at the end of the Removal era (1850). During the subsequent Reservation era (1850 - 1887), the number of operating boarding schools increased from 23 to 173. The number of operating facilities continued to increase during the Assimilation era to a maximum of 242 in 1908, followed by a decline to 174 facilities recorded as operating in 1930, a similar level of facilities operating as at the beginning of the period.

@@

Figure \@ref(fig:bsmap) shows the geographic distribution of boarding schools in operation when enrollments were at a 1830 - 1930 period maximum (242 facilities documented as operating in 1907). The Department of the Interior has identified Indian Boarding schools operating in 1907 in most regions of the country, with the exception of New England and the Southeast. There were at least 64 boarding school facilities operating in Oklahama in 1907, by far the highest number in the US. There were 8 states with at least 10 facilities in operation in 1907: South Dakota (24), Minnesota (18), Arizona (14), Montana (14), New Mexico (11), Washington (11), Alaska (10, not displayed), and Wisconsin (10). There were at least 5 boarding schools in operation in 7 states: California (7), North Dakota (7), Oregon (7), Colorado (5), Hawaii (5, not displayed), Idaho (5), and Kansas (5). States with at least one confirmed facility in operation in 1907 included Nebraska (3), Nevada (3), Pennsylvania (3), Utah (3), Wyoming (3), Iowa (2), Michigan (2), Arkansas (1), Illinois (1), Missouri (1), New York (1), North Carolina (1), and Virginia (1). The data do not record facilities operating in 1907 in the remaining 21 states.

```{r bsmap, fig.cap = "Boarding schools in operation in 1907 by state"}
bs_ts<-bs %>% 
  left_join(st)
ggplot(bs_ts %>% 
         filter(year == 1907),
       aes(fill = boarding_schools,
           geometry = geometry)) + 
  geom_sf() + 
  scale_fill_viridis_c(trans = "sqrt") + 
  theme_void() + 
  labs(fill = "Facilities\noperating") + 
  theme(legend.position = "bottom")
```

@@

We describe the results of our regression models in terms of expected values of each infrastructure outcome described above. We compute mean expected values and 95 percent credible intervals for each decennial year 1900 - 1930 from model posterior distributions for each outcome across three scenarios that closely correspond to the regional geography of the US during this period. We describe these scenarios in Table \@ref(tab:scen). All models include state-specific intercepts and slopes over time to account for other unmeasured features of states that impact both levels of employment and rates of change in employment.

The "High land redistribution" scenario is intended to describe states with a high volume of lands patented under the Morill and Homestead acts. These states typically had relatively low levels of Black tenant farmers per capita and white urban immigrants per capita (see Figure \@ref(fig:three-maps). For this scenario, we fix the sum of cumulative Homestead Act and Morill Act acres at the 90th percentile of the observed data (12.5 million acres), Black tenant farm residents per capita at the 5th percentile of the observed data (about 2 per 100,000 population), and white urban immigrants per capita at the 5th percentile of the observed data (about 2.6 per 1,000 population). This scenario is similar to the observed values of North Dakota, South Dakota, Montana, Colorado, Wyoming, and Kansas.

The "High sharecropping scenario" is intended to describe states with a high volume of Black residents on tenant farms. These states, all in the US South, typically had very small urban populations and did not see a high volume of lands parceled under the Homestead and Morill acts. For this scenario, we fix Black tenant farm residents per capita at the 90th percentile of the observed data (143 per 1,000 population), white urban immigrant population per capita at the 5th percentile of the observed data (2.6 per 1,000 population), and the sum of Homestead and Morill acres patented at the 5th percentile of the observed data (0 acres). This scenario is most similar to Georgia and South Carolina during this period. The scenario is also similar to Louisiana, Alabama and Mississippi, though these states did have substantial acres of land patented under the Homestead and Morill acts prior to 1900.

```{r scen, results = "asis", tab.cap= "Scenarios specified for model predictions"}
year_c = 0:3
STUSPS <- "XX"
tot_pop = 1e3

pred_dat<-data.frame(
  scen = c("High sharecropping", "High urban immigrants", "High land redistribution"),
  blk_sharecrop_pc = quantile(mdat$blk_sharecrop_pc, c(0.9, 0.05, 0.05)),
  urban_immig_pc = quantile(mdat$urban_immig_pc, c(0.05, 0.9, 0.05)),
  homestead_plus_morill = quantile(mdat$homestead_plus_morill, c(0.05, 0.05, 0.9))) %>% 
  expand_grid(year_c, tot_pop)

tab_dat<-pred_dat %>% 
  filter(year_c == 0) %>% 
  distinct() %>% 
  mutate(blk_sharecrop_pc = blk_sharecrop_pc * 1e3,
         urban_immig_pc = urban_immig_pc * 1e3,
         homestead_plus_morill = homestead_plus_morill / 1e6) %>% 
  select(-year_c, -tot_pop) %>% 
  rename("Scenario" = scen,
         "Black tenant\nfarmers per 1,000 pop." = blk_sharecrop_pc,
         "White urban\nimmigrants per 1,000 pop." = urban_immig_pc,
         "Homestead + Morill\nacres patented" = homestead_plus_morill)

kable(tab_dat, digits = 2)
```

The "High urban immigrants" scenario corresponds to state with large urban industrial centers that experienced a high volume of European immigration between 1900 and 1930. These states typically had very low numbers of Black tenant farmers during this period. States in the Northeast did not see lands patented to settlers under the Homestead and Morill acts. However, states on the West Coast and Midwest, some of which had large urban immigrant populations, were exposed to high-levels of homesteading and settler land redistribution during this period. For ease of comparison, we set the "High urban immigrant" scenario at the 90th percentile of observed white urban immigrants per capita (176 per 1,000 residents), and the 5th percentile of observed Black tenant farm residents per capita and cumulative Homestead and Morill acres patented. This scenario is similar to conditions in Connecticut, Illinois, Massachusetts, New Jersey, New York, and Rhode Island.

In Figure \@ref(fig:emp-m-vis) we display our estimates of the spatial and temporal relationships between employment in key social policy professions and settler colonial project indicators. In Figure \@ref(fig:gq-m-vis) we show our estimates for the relationship between institutional group quarters populations and settler colonial project indicators. And in Figure \@ref(fig:bs-m-mvs) we show our estimates of the spatial and temporal relationships between operating Federal Indian boarding schools and these measures of settler colonial projects.

@@

### Employment

There was a dramatic increase in employment in education, policing, and social welfare between 1900 and 1930 nationally (see Figure \@ref(fig:emp-ts)), with substantial differences across regions of the US (see Figure \@ref(fig:emp-maps)). We construct a series of regression models to explore how the geography of settler projects related to both levels of employment and growth in employment across states. We display key results from these models as posterior expected values for focal scenarios in Figure \@ref(fig:emp-m-vis). We find that there was substantial growth in employment across all settler project contexts between 1900 and 1930, initial levels of employment and rates of change in employment varied dramatically across contexts.

```{r emp-m-vis, fig.cap = "Posterior expected employment rates per 1,000 population by employment type across settler project contexts, US States 1900 - 1930. Mean posterior expected value (line) and 95 percent credible interval (shaded)."}
preds_teach<-teachers_m1 %>% 
  add_epred_draws(re_formula = NA,
                  newdata = pred_dat) %>% 
  mutate(type = "Teachers")

preds_cops<-cops_m1 %>% 
  add_epred_draws(re_formula = NA,
                  newdata = pred_dat)%>% 
  mutate(type = "Police")

preds_socwork<-socwork_m1 %>% 
  add_epred_draws(re_formula = NA,
                  newdata = pred_dat)%>% 
  mutate(type = "Social workers")

p1<-ggplot(preds_teach,
       aes(x = year_c*10 + 1900, y = .epred)) + 
  stat_lineribbon(show.legend = F, 
                  .width = 0.95, 
                  fill = "gray50",
                  linewidth = 0.5) +
  theme_bw() + 
  theme(panel.spacing = unit(0.4, "cm")) +
  facet_wrap(~scen) + 
  labs(y = "",
       x = "",
       subtitle = "1. Teachers")

p2<-ggplot(preds_cops,
       aes(x = year_c*10 + 1900, y = .epred)) + 
  stat_lineribbon(show.legend = F, 
                  .width = 0.95, 
                  fill = "gray50",
                  linewidth = 0.5) +
  theme_bw() + 
  theme(panel.spacing = unit(0.4, "cm")) +
  facet_wrap(~scen) + 
  labs(y = "Employment rate",
       x = "",
       subtitle = "2. Police")
